---
title: "Redlining Impacts on Environmental Justice"
author: "Kat Le"
output: html_document
date: "2023-10-17"  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in the past. The United States has a long history of racial segregation which is still visible. During the 1930's the Home Owners' Loan Corporation (HOLC), as part of the New Deal, rated neighborhoods based on their perceived safety for real estate investment. Their ranking system, (A (green), B (blue), C (yellow), D (red)) was then used to block access to loans for home ownership. Colloquially known as "redlining", this practice has had widely-documented consequences not only for community wealth, but also health.[^1] Redlined neighborhoods have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship between institutional and individual racial discrimination and health status. American journal of public health, 98(Supplement_1), S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A. (2021). Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States. Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the environments communities are exposed to, it has also shaped our observations of biodiversity.[^4] Community or citizen science, whereby individuals share observations of species, is generating an enormous volume of data. Ellis-Soto and co-authors found that redlined neighborhoods remain the most undersampled areas across 195 US cities. This gap is highly concerning, because conservation decisions are made based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. Nature Human Behaviour, 1-9.

Check out coverage by [EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

We will be working with data from the United States Environmental Protection Agency's EJScreen: Environmental Justice Screening and Mapping Tool.

According to the [US EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents or other stakeholders as they search for environmental or demographic information. It can also support a wide range of research and policy goals. The public has used EJScreen in many different locations and in many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice in our work,\
> - to assist our stakeholders in making informed decisions about pursuing environmental justice and,\
> - to create a common starting point between the agency and the public when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and [block group](https://en.wikipedia.org/wiki/Census_block_group) levels. You will be working with block group data that has been downloaded from the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To understand the associated data columns, you will need to explore the Technical Documentation and column description spreadsheet available in the data folder. I also encourage you to explore the [limitations and caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen) of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship Lab](https://dsl.richmond.edu/) at the University of Richmond have digitized maps and information from the HOLC as part of the [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) project.

We will be working with maps of HOLC grade designations for Los Angeles. Information on the data can be found [here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., "Mapping Inequality," American Panorama, ed. Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023, <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest aggregator of biodiversity observations in the world. Observations typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

## Assignment

### Investigate the legacy of redlining in current environmental (in)justice

Load relevant packages.

```{r include=TRUE, warning=FALSE, message=FALSE, results = FALSE}
library(tidyverse)
library(sf)
library(here)
library(tmap)
library(ggspatial)
library(ggsn)
library(plotly)
library(gt)
```

Read in EJScreen data and filter to Los Angeles County (5 points)

```{r include=TRUE, warning=FALSE, message=FALSE, results = FALSE}
la_data <- st_read(here("data", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) %>% 
  filter(CNTY_NAME == "Los Angeles County")
```

Make a map of wastewater discharge by census block groups. Indicate which census block groups are above the 95th percentile of national values for wastewater discharge by adding a centroid. (10 points)

```{r include=TRUE, warning=FALSE, message=FALSE, results = FALSE}
wastewater_above95 <- la_data %>% 
  filter(P_PWDIS > 95)


map1 <- ggplot() +
  geom_sf(data = la_data, aes(fill = PWDIS)) +
  geom_sf(data = st_centroid(wastewater_above95),
          aes(value = P_PWDIS), color = "red") +
  labs(title = "LA Waste Water Discharge Percentile (PWDIS)") +
    annotation_scale(data = la_data, location = "br") +
    theme_linedraw() +
    annotation_north_arrow(location = "bl",
                           pad_x = unit(0.02, "in"),
                           pad_y = unit(0.02, "in"),
                           style = ggspatial::north_arrow_nautical(fill = c("grey40", "white")))
ggplotly(map1)
```

Find the percent of census block groups that have:\
- less than 5% of the population is considered low income (5 points)

```{r }
object1 <- la_data %>% 
  filter(LOWINCPCT < 0.05)

answer1 <- (nrow(object1)/nrow(la_data))*100
answer1
```

Find the percent of census block groups that are:\
- above the 80th percentile for Particulate Matter 2.5 AND\
- above the 80th percentile for Superfund proximity (10 points)

```{r }
object2 <- la_data %>%  
  filter(P_PM25 > 80 & P_PNPL > 80)

answer2 <- (nrow(object2)/nrow(la_data))*100
answer2
```

Import redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid()

LA_county <- st_read(here("data/los-angeles-county.geojson")) %>%
  st_make_valid() # county layer w same CRS as LA_redlining
```

Make a map of historical redlining boundaries, colored by HOLC grade. (5 points)

```{r }
map2 <- ggplot() +
  geom_sf(data = LA_county) +
  geom_sf(data = LA_redlining, aes(fill = grade)) +
  coord_sf(expand = TRUE) +
    annotation_scale(data = LA_county, location = "br") +
    theme_linedraw() +
  labs(title = "Historical redlining in LA by HOLC Grade") +
    annotation_north_arrow(location = "tr",
                           pad_x = unit(0.05, "in"),
                           pad_y = unit(0.05, "in"),
                           style = ggspatial::north_arrow_nautical(fill = c("grey40", "white")))


ggplotly(map2)
```

Find the number of census block groups that fall within areas with HOLC grades hint: make sure the CRS match (15 points)

```{r }
la_data_4326 <- st_transform(la_data, crs = "EPSG: 4326")

st_crs(la_data_4326)
st_crs(LA_redlining)

la_redlining_sgbp <- st_intersects(x = la_data_4326, y = LA_redlining)

la_redlining_logical <- lengths(la_redlining_sgbp) > 0

LA_redlining_joined <- la_data_4326[la_redlining_logical,]

answer3 <- nrow(LA_redlining_joined)
answer3

```

Summarize current conditions based on EJScreen data within historical redlining categories using the mean of the following variables:\
-% low income.\
- percentile for particulate Matter 2.5.\
- percentile for low life expectancy.\
- percentile for air toxics cancer risk (20 points)

```{r }
summarize_la <- st_join(x = la_data_4326, y = LA_redlining) %>% 
  group_by(grade) %>% 
  filter(!is.na(grade)) %>% 
  summarize("LowIncPct Mean" = mean(LOWINCPCT, na.rm = TRUE),
            "P_PM25 Mean" = mean(P_PM25, na.rm = TRUE),
            "P_LifeExpPct Mean" = mean(P_LIFEEXPPCT, na.rm = TRUE),
            "P_Cancer Mean" = mean(P_CANCER, na.rm = TRUE),
            "% POC" = mean(PEOPCOLORPCT, na.rm = TRUE))

answer4 <- gt(st_drop_geometry(summarize_la)) %>% 
  tab_header(
    title = "Environmental Metrics of Redlining Boundaries by HOLC Grade"
  )
answer4


```

Please write 2-3 sentences reflecting on these results. (5 points)

Answer: The effects of redlining can still be seen today as evidenced by the environmental data above. In areas that were deemed more hazardous by the government and real estate agencies, the quality of life is significantly lower (higher % of people are low income, more air pollution, lower life expectancy, and higher risk of air toxic cancer). I added a column for percentage of people of color to the summary table to show how race is an important consideration in determining which districts get more investments from the government + real estate agencies.

### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC grads, find the percent of observations within each redlining categories and plot results. hint: make sure that the bird observations have the same CRS as redlining data. (20 points)

```{r}

birds <- st_read(here("data/gbif-birds-LA/gbif-birds-LA.shp")) %>%
 filter(year == 2022) # since question says from year 2022

st_crs(birds)
st_crs(LA_redlining)

birds_redlining_joined <- st_join(x = birds, y = LA_redlining, join = st_within)

bird_byholc <- birds_redlining_joined %>% 
    group_by(grade) %>% 
    drop_na(grade) %>% 
    summarize(bird_count = n()) %>% 
    mutate(bird_total = sum(bird_count)) %>% 
    mutate(bird_percent = (bird_count / bird_total)*100)


answer5 <- ggplot(data = bird_byholc, 
                  aes(x = grade, y = bird_percent, fill = grade)) +
  geom_col() +
  labs(title = "Percentage of Birds within LA Redlining boundaries by HOLC Grade",
       y = "Bird %",
       x = "HOLC Grade")
answer5
```

Please write 2-3 sentences reflecting on these results. Are they surprising? If so, why? What are some potential implications? (5 points)

Answer:

Given that this analysis is simply on bird observations and not on species diversity, these results are not surprising to me because I grew up in LA and have seen flocks of birds work together to take off commercial trash can lids so that they can sift through people's trash for food. In areas that are potentially more populated per area & produce more trash like areas C & D, it makes sense to me that there are more birds.
